---
title: "Parameterizing Distributions for PSA"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Parameterizing Distributions for PSA}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

## Why PSA Needs the Right Distribution

Probabilistic Sensitivity Analysis (PSA) requires sampling model parameters from appropriate distributions. Using the wrong distribution can produce impossible values (e.g., probabilities > 1 or negative costs) and bias the results.

| Parameter Type | Recommended Distribution | Reason |
|---|---|---|
| Probabilities / Utilities | Beta | Bounded between 0 and 1 |
| Costs / Resource Use | Gamma | Non-negative, right-skewed |
| Hazard Ratios / Odds Ratios | LogNormal | Non-negative, multiplicative |

## Tutorial 1: Utilities with Beta Distribution

### The Scenario — EQ-5D Utility in COPD

A quality-of-life study in COPD (Rutten-van Molken et al., Chest 2006) reports the EQ-5D utility for "Moderate COPD (GOLD Stage II)" as:

> Mean = **0.76**, Standard Error = **0.03**

### The Method of Moments

$$\alpha = \mu \left(\frac{\mu(1-\mu)}{SE^2} - 1\right)$$

$$\beta = (1-\mu) \left(\frac{\mu(1-\mu)}{SE^2} - 1\right)$$

### Worked Example

```{r}
mu <- 0.76
se <- 0.03

common <- (mu * (1 - mu)) / se^2 - 1
alpha <- mu * common
beta_param <- (1 - mu) * common

cat("Alpha:", round(alpha, 2), "\n")
cat("Beta:", round(beta_param, 2), "\n")

# Verify: sample and check
set.seed(42)
samples <- rbeta(10000, alpha, beta_param)
cat("\nVerification (10,000 samples):\n")
cat("Mean:", round(mean(samples), 4), "(target:", mu, ")\n")
cat("SE:", round(sd(samples), 4), "(target:", se, ")\n")
cat("Range:", round(range(samples), 4), "\n")
```

All values are bounded between 0 and 1, as required for utilities.

## Tutorial 2: Costs with Gamma Distribution

### The Scenario — Surgery Cost (AIIMS Costing Study)

A micro-costing study reports the cost of CABG surgery as:

> Mean = **INR 2,50,000**, Standard Error = **INR 50,000**

### The Method of Moments

$$k = \frac{\mu^2}{SE^2}, \quad \theta = \frac{SE^2}{\mu}$$

### Worked Example

```{r}
mu_cost <- 250000
se_cost <- 50000

k <- mu_cost^2 / se_cost^2
theta <- se_cost^2 / mu_cost

cat("Shape (k):", round(k, 2), "\n")
cat("Scale (theta):", round(theta, 2), "\n")

# Verify
set.seed(42)
samples_cost <- rgamma(10000, shape = k, scale = theta)
cat("\nVerification (10,000 samples):\n")
cat("Mean: INR", format(round(mean(samples_cost)), big.mark = ","), "(target:", format(mu_cost, big.mark = ","), ")\n")
cat("SE: INR", format(round(sd(samples_cost)), big.mark = ","), "(target:", format(se_cost, big.mark = ","), ")\n")
cat("Min: INR", format(round(min(samples_cost)), big.mark = ","), "(always positive)\n")
```

## Tutorial 3: Hazard Ratios with LogNormal Distribution

### The Scenario — HR from Network Meta-Analysis

A network meta-analysis reports the HR for Drug A vs placebo as:

> HR = **0.72**, 95% CI: **0.58 to 0.89**

### The Method

```{r}
hr_mean <- 0.72
hr_low <- 0.58
hr_high <- 0.89

mu_log <- log(hr_mean)
se_log <- (log(hr_high) - log(hr_low)) / (2 * 1.96)
sigma2_log <- se_log^2

cat("mu_log:", round(mu_log, 4), "\n")
cat("sigma_log:", round(se_log, 4), "\n")

# Verify
set.seed(42)
samples_hr <- rlnorm(10000, meanlog = mu_log, sdlog = se_log)
cat("\nVerification (10,000 samples):\n")
cat("Median HR:", round(median(samples_hr), 3), "(target:", hr_mean, ")\n")
cat("2.5th percentile:", round(quantile(samples_hr, 0.025), 3), "(target:", hr_low, ")\n")
cat("97.5th percentile:", round(quantile(samples_hr, 0.975), 3), "(target:", hr_high, ")\n")
```

## When SE is Missing: The Rule of 4

If only a range (min, max) is available instead of SE, estimate:

$$SE \approx \frac{High - Low}{4}$$

This assumes the range covers approximately 95% of the distribution (i.e., ±2 SE).

```{r}
# Example: cost range INR 1,50,000 to INR 3,50,000
cost_low <- 150000
cost_high <- 350000
se_approx <- (cost_high - cost_low) / 4

cat("Estimated SE:", format(se_approx, big.mark = ","), "\n")
```

## References

- Briggs A, Claxton K, Sculpher M. *Decision Modelling for Health Economic Evaluation*. Oxford University Press; 2006. Chapter 4.
- Rutten-van Molken MP, et al. Modelling the 5-year cost effectiveness of tiotropium. *Chest*. 2006;130(5):1549.
